---
title: "NBA Scoring Analysis"
author: "Varun Gullanki, Anirudh Ganesan, Zezhou Zhuang"
format: pdf
editor: visual
---

## Introduction

In professional basketball, player performance is influenced by a range of physical and experiential factors. This project explores how **height** and **age** relate to **scoring performance** in the NBA. We investigate the following questions:

1.  Do taller players score more points?
2.  At what age do players typically reach their scoring peak?
3.  What combination of height and age is optimal for scoring?

These insights can support coaches, analysts, and scouts in evaluating player potential and optimizing team composition.

## Data Provenance

The dataset was compiled and uploaded by Justinas to Kaggle. It contains statistics from multiple NBA seasons and includes variables such as player height, age, and points per game.Data source: Justinas. “NBA Players Data.” *Kaggle*, 2023, <https://www.kaggle.com/datasets/justinas/nba-players-data>. Accessed May 2025.

## FAIR/CARE Principles

-   **Findable**: Data is publicly available on a trusted open data platform.
-   **Accessible**: Provided in a machine-readable format (CSV).
-   **Interoperable**: Compatible with common tools such as R and Python.
-   **Reusable**: Contains structured and labeled variables that allow for versatile analysis.

## Summary Statistics

To prepare the data for analysis, we first converted player height from centimeters to inches. Then we filtered out missing or zero point values to maintain analytical accuracy.

Below is a summary of the cleaned dataset:

```{r echo=FALSE}
#| message: false
#| warning: false
#| paged-print: false
library(tidyverse)
nba <- read_csv("/Users/varun/Downloads/all_seasons.csv")
nba_clean <- nba %>%
  mutate(player_height = player_height * 0.393701) %>%
  filter(!is.na(pts), !is.na(age), !is.na(player_height), pts > 0)

# Select key variables for summary
summary_data <- nba_clean %>%
  select(age, player_height, pts)

knitr::kable(summary(summary_data), caption = "Summary Statistics for Age, Height (in), and Points")
```


## Q1: Does Player Height Affect Scoring?

```{r echo=FALSE}
#| warning: false
ggplot(nba_clean, aes(x = player_height, y = pts)) +
  geom_point(alpha = 0.3, color = "darkgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  labs(
    title = "Player Height vs. Points Per Game",
    x = "Player Height (inches)",
    y = "Points Per Game"
  )
```

**Interpretation**:\
The scatterplot reveals a broad range of scoring outputs across different heights. While there is a slight upward trend, the weak correlation suggests that **height is not a strong predictor of scoring**. Role-specific factors (e.g., guards handling the ball more) may have a greater impact than height alone.


## Q2: At What Age Do NBA Players Peak in Scoring?

```{r echo=FALSE}
#| warning: false
age_avg_pts <- nba_clean %>%
  group_by(age) %>%
  summarize(mean_pts = mean(pts), count = n()) %>%
  filter(count > 20)

ggplot(age_avg_pts, aes(x = age, y = mean_pts)) +
  geom_line(color = "firebrick", size = 1.3) +
  geom_point(color = "black") +
  labs(
    title = "Average Points Per Game by Age",
    x = "Age",
    y = "Average Points Per Game"
  )
```

**Interpretation**:\
NBA players tend to improve their scoring output through their early 20s, peaking between **ages 26–28**. After this range, scoring declines — likely due to factors such as reduced athleticism, injuries, or changes in team role.


## Q3: What Is the Optimal Height/Age for Scoring?

```{r echo=FALSE}
nba_grid <- nba_clean %>%
  group_by(age, player_height) %>%
  summarize(mean_pts = mean(pts), .groups = "drop")

ggplot(nba_grid, aes(x = player_height, y = age, fill = mean_pts)) +
  geom_tile() +
  labs(
    title = "Optimal Height and Age for Scoring",
    x = "Player Height (inches)",
    y = "Age",
    fill = "Avg Points"
  )
```

**Interpretation**:\
This heatmap shows that players between **75–79 inches** (roughly 6'3" to 6'7") and **ages 25–28** tend to average the most points per game. This region represents athletic and experienced guards and forwards — a valuable profile for high-scoring performance.


## Data Visualization Summary

We created three types of visualizations: - A **scatterplot** to analyze the height-to-scoring relationship. - A **line graph** to observe scoring trends across ages. - A **heatmap** to explore the intersection of height and age with scoring output.

These plots collectively show that **experience and playing role outweigh raw height** when it comes to scoring in the NBA.

## Conclusion

-   **Height is not a major driver of scoring**. Players of all sizes can be high scorers depending on their role.
-   **Age has a clear peak window** around 26–28 years, reflecting the balance of skill and physical condition.
-   The best scoring performances occur in players with a combination of **moderate-to-tall height and mid-to-late 20s age**.

This analysis emphasizes that player performance depends on multiple factors — not just physical measurements. Future research could control for player position, minutes played, or usage rate to refine these insights.

## Code Appendix

```{r codeAppend, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```
